% !TEX root = ../master_thesis.tex
\chapter{Background and Related Work}
\label{chap:background}

\section{Research Gap}
\label{sec:gap}

The preceding sections have reviewed GUI quality assessment, LLM-generated
interfaces, LLM-as-a-judge methodology as an evaluation tool, and
multimodal web agents.
Table~\ref{tab:related-work} places the most relevant prior works along
seven dimensions that together characterise a comprehensive GUI quality
evaluation framework.
Taken together, three observations motivate the present thesis.

\paragraph{Existing website-generation benchmarks do not directly answer generator-leaderboard questions.}
UIClip~\cite{uiclip} demonstrated that vision-language models can perform
pairwise UI comparison, but it relies on models that predate the current
generation of multimodal LLMs and limits evaluation to a binary preference.
The LLM-as-a-judge literature~\cite{llm_judge,li-etal-2025-generation}
has established scalable evaluation methods for text, yet those methods
also introduce known bias and calibration concerns. In this thesis they
are therefore treated as practical scoring instruments rather than as the
main leaderboard target.
Vision2Web~\cite{vision2web} moves closer to this thesis by evaluating
visual website development with both visual judging and agent-based
verification, but its primary object is coding-agent website development
rather than a thesis-specific, multi-metric leaderboard for comparing
generated GUI submissions across static and dynamic metrics.

\paragraph{Requirement fidelity and rubric-based assessment are largely absent from GUI evaluation.}
Design2Code~\cite{design2code} evaluates how faithfully generated code
reproduces a visual reference, but does not assess richer quality dimensions
such as visual hierarchy, information clarity, or perceived usability.
The Designer Feedback study~\cite{designer_feedback} found that simple
pairwise rankings among evaluators agreed at only around 50\%, suggesting
that coarse binary judgements are insufficient for reliable GUI assessment.
A structured rubric grounded in established usability
theory~\cite{nielsen1990heuristic,iso9241-11} can provide finer-grained and
more reproducible quality signals.

\paragraph{Static visual metrics have rarely been related to dynamic interaction outcomes.}
VisualWebArena~\cite{visualwebarena} and WebArena~\cite{webarena}
evaluate multimodal agents on goal-directed web tasks, but do not assess
the GUI quality of the interfaces themselves.
How static quality scores relate to task completion, interaction feedback,
and functional behaviour remains an open empirical question.
Answering this question requires a benchmark that keeps generated interfaces
executable so that static and dynamic evaluations can be performed on the
same set of items.

The present thesis addresses all three gaps simultaneously by combining
a two-track benchmark, a rubric-based scoring protocol, controlled
generation metadata, validity gates, and a static-plus-dynamic leaderboard
for comparing the quality of LLM-generated GUIs. Judge-model bias and
human-label subjectivity are acknowledged as limitations of the scoring
process, not as the central object of the leaderboard.

\begin{landscape}
\begin{table}[p]
  \centering
  \caption{Comparison of related work across seven evaluation dimensions.
           \checkmark~= fully addressed;
           $\triangle$~= partially addressed;
           $\times$~= not addressed;
           ---~= not applicable to this line of work.}
  \label{tab:related-work}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{p{3.8cm}ccccccc}
    \toprule
    \textbf{Work}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Multi-}\\\textbf{dimensional}\\\textbf{rubric}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Generated}\\\textbf{GUI}\\\textbf{quality}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Bias and}\\\textbf{reliability}\\\textbf{analysis}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Human}\\\textbf{reference}\\\textbf{labels}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Requirement}\\\textbf{fidelity}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Dynamic}\\\textbf{interaction}\\\textbf{eval}\end{tabular}
      & \begin{tabular}[c]{@{}c@{}}\textbf{Benchmark}\\\textbf{artifact}\end{tabular} \\
    \midrule
    UIClip~\cite{uiclip}
      & $\times$ & $\times$ & $\times$ & \checkmark & $\times$ & $\times$ & $\times$ \\
    Designer Feedback~\cite{designer_feedback}
      & $\times$ & \checkmark & $\times$ & \checkmark & $\triangle$ & $\times$ & $\times$ \\
    Design2Code~\cite{design2code}
      & $\triangle$ & \checkmark & $\times$ & \checkmark & \checkmark & $\times$ & \checkmark \\
    Vision2Web~\cite{vision2web}
      & $\triangle$ & \checkmark & $\triangle$ & $\triangle$ & \checkmark & \checkmark & \checkmark \\
    MT-Bench / Chatbot Arena~\cite{llm_judge}
      & $\triangle$ & --- & $\triangle$ & \checkmark & --- & --- & $\triangle$ \\
    LLM-as-a-Judge survey~\cite{li-etal-2025-generation}
      & \checkmark & --- & \checkmark & --- & --- & --- & $\times$ \\
    Position Bias Study~\cite{shi-etal-2025-judging}
      & $\times$ & --- & \checkmark & \checkmark & --- & --- & $\times$ \\
    WebArena~\cite{webarena}
      & $\times$ & $\triangle$ & $\times$ & \checkmark & $\triangle$ & \checkmark & \checkmark \\
    VisualWebArena~\cite{visualwebarena}
      & $\times$ & \checkmark & $\times$ & \checkmark & $\triangle$ & \checkmark & \checkmark \\
    \midrule
    \textbf{This thesis}
      & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    \bottomrule
  \end{tabular}
\end{table}
\end{landscape}
